---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 150
)
```

# ggbumpribbon

<!-- badges: start -->
[![R-CMD-check](https://github.com/sondreskarsten/ggbumpribbon/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sondreskarsten/ggbumpribbon/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Sigmoid-curved filled ribbons for rank comparison charts in ggplot2.

```{r reputation, echo = FALSE, fig.width = 7, fig.height = 17, dpi = 200}
library(ggplot2)
library(ggbumpribbon)
library(ggflags)
library(countrycode)

ranks <- data.frame(stringsAsFactors = FALSE,
  country   = c("Switzerland","Norway","Sweden","Canada","Denmark","New Zealand","Finland",
                "Australia","Ireland","Netherlands","Austria","Japan","Spain","Italy","Belgium",
                "Portugal","Greece","UK","Singapore","France","Germany","Czechia","Thailand",
                "Poland","South Korea","Malaysia","Indonesia","Peru","Brazil","U.S.","Ukraine",
                "Philippines","Morocco","Chile","Hungary","Argentina","Vietnam","Egypt","UAE",
                "South Africa","Mexico","Romania","India","Turkey","Qatar","Algeria","Ethiopia",
                "Colombia","Kazakhstan","Nigeria","Bangladesh","Israel","Saudi Arabia","Pakistan",
                "China","Iran","Iraq","Russia"),
  rank_from = c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24,25,26,27,28,
                29,30,31,32,33,34,35,36,37,38,39,40,41,42,43,44,45,51,47,49,50,52,53,54,55,56,
                57,58,59,60),
  rank_to   = c(1,3,4,2,6,7,5,11,10,9,12,8,14,13,17,15,16,18,19,21,20,25,24,23,31,29,34,27,
                28,48,26,33,30,35,32,38,37,36,40,42,39,41,45,43,44,46,51,50,49,52,54,55,53,56,
                57,59,58,60))

exit_only  <- data.frame(country = c("Cuba","Venezuela"),  rank_from = c(46,48), stringsAsFactors = FALSE)
enter_only <- data.frame(country = c("Taiwan","Kuwait"),   rank_to   = c(22,47), stringsAsFactors = FALSE)

ov <- c("U.S."="us","UK"="gb","South Korea"="kr","Czechia"="cz","Taiwan"="tw","UAE"="ae")
iso <- function(x) ifelse(x %in% names(ov), ov[x],
  tolower(countrycode(x, "country.name", "iso2c", warn = FALSE)))

ranks$iso2      <- iso(ranks$country)
exit_only$iso2  <- iso(exit_only$country)
enter_only$iso2 <- iso(enter_only$country)

ranks_long <- data.frame(
  x       = rep(1:2, each = nrow(ranks)),
  y       = c(ranks$rank_from, ranks$rank_to),
  group   = rep(ranks$country, 2),
  country = rep(ranks$country, 2),
  iso2    = rep(ranks$iso2, 2))

lbl_l <- ranks_long[ranks_long$x == 1, ]
lbl_r <- ranks_long[ranks_long$x == 2, ]

ggplot(ranks_long, aes(x, y, group = group, fill = after_stat(avg_y))) +
  geom_bump_ribbon(alpha = 0.85, width = 0.8) +
  scale_fill_gradientn(
    colours = c("#2ecc71","#a8e063","#f7dc6f","#f0932b","#eb4d4b","#c0392b"),
    guide = "none") +
  scale_y_reverse(expand = expansion(mult = c(0.015, 0.015))) +
  scale_x_continuous(limits = c(0.15, 2.85)) +
  geom_text(data = lbl_l, aes(x = 0.94, y = y, label = y),
            inherit.aes = FALSE, hjust = 1, colour = "white", size = 2.2) +
  geom_flag(data = lbl_l, aes(x = 0.88, y = y, country = iso2),
            inherit.aes = FALSE, size = 3) +
  geom_text(data = lbl_l, aes(x = 0.82, y = y, label = country),
            inherit.aes = FALSE, hjust = 1, colour = "white", size = 2.2) +
  geom_text(data = lbl_r, aes(x = 2.06, y = y, label = y),
            inherit.aes = FALSE, hjust = 0, colour = "white", size = 2.2) +
  geom_flag(data = lbl_r, aes(x = 2.12, y = y, country = iso2),
            inherit.aes = FALSE, size = 3) +
  geom_text(data = lbl_r, aes(x = 2.18, y = y, label = country),
            inherit.aes = FALSE, hjust = 0, colour = "white", size = 2.2) +
  geom_text(data = exit_only, aes(x = 0.94, y = rank_from, label = rank_from),
            inherit.aes = FALSE, hjust = 1, colour = "grey55", size = 2.2) +
  geom_flag(data = exit_only, aes(x = 0.88, y = rank_from, country = iso2),
            inherit.aes = FALSE, size = 3) +
  geom_text(data = exit_only, aes(x = 0.82, y = rank_from, label = country),
            inherit.aes = FALSE, hjust = 1, colour = "grey55", size = 2.2) +
  geom_text(data = enter_only, aes(x = 2.06, y = rank_to, label = rank_to),
            inherit.aes = FALSE, hjust = 0, colour = "grey55", size = 2.2) +
  geom_flag(data = enter_only, aes(x = 2.12, y = rank_to, country = iso2),
            inherit.aes = FALSE, size = 3) +
  geom_text(data = enter_only, aes(x = 2.18, y = rank_to, label = country),
            inherit.aes = FALSE, hjust = 0, colour = "grey55", size = 2.2) +
  annotate("text", x = 1, y = -1.5, label = "2024 Rank",
           colour = "white", size = 4.5, fontface = "bold") +
  annotate("text", x = 2, y = -1.5, label = "2025 Rank",
           colour = "white", size = 4.5, fontface = "bold") +
  labs(title    = "COUNTRIES WITH THE BEST REPUTATIONS IN 2025",
       subtitle = "Reputation Lab ranked the reputations of 60 leading economies\nin 2025, shedding light on their international standing.",
       caption  = "Source: Reputation Lab | Made with ggbumpribbon") +
  theme_bump()
```

<details>
<summary>Code to reproduce (requires ggflags + countrycode)</summary>

```{r ref.label = "reputation", eval = FALSE}
```

Flags require [ggflags](https://github.com/jimjam-slam/ggflags):
`install.packages("ggflags", repos = c("https://jimjam-slam.r-universe.dev", "https://cloud.r-project.org"))`

</details>

## The gap

| Package | What it does | What it lacks |
|---------|-------------|---------------|
| **ggbump** | Sigmoid *lines* via `geom_bump()` | No filled ribbons |
| **ggforce** | Bezier *ribbons* via `geom_diagonal_wide()` | Bezier, not sigmoid curve shape |
| **ggsankey** | Sankey-style ribbon bumps | *Stacked* positioning, not rank-positioned. GitHub-only |
| **ggbumpribbon** | **Sigmoid filled ribbons at exact rank positions** | — |

## Installation

```{r installation, eval = FALSE}
# install.packages("pak")
pak::pak("sondreskarsten/ggbumpribbon")
```

## Minimal example

```{r basic, fig.width = 6, fig.height = 4}
library(ggplot2)
library(ggbumpribbon)

df <- data.frame(
  x     = rep(1:2, each = 5),
  y     = c(1, 2, 3, 4, 5, 3, 1, 5, 2, 4),
  group = rep(LETTERS[1:5], 2)
)

lbl_l <- df[df$x == 1, ]; lbl_r <- df[df$x == 2, ]

ggplot(df, aes(x, y, group = group, fill = after_stat(avg_y))) +
  geom_bump_ribbon(alpha = 0.85) +
  scale_fill_gradientn(colours = c("#2ecc71", "#f7dc6f", "#eb4d4b"), guide = "none") +
  scale_y_reverse() +
  scale_x_continuous(limits = c(0.3, 2.7)) +
  geom_text(data = lbl_l, aes(x = 0.92, y = y, label = paste(y, group)),
            inherit.aes = FALSE, hjust = 1, size = 4) +
  geom_text(data = lbl_r, aes(x = 2.08, y = y, label = paste(group, y)),
            inherit.aes = FALSE, hjust = 0, size = 4) +
  theme_void()
```

## Multi-period

Ribbons chain automatically across 3+ time points:

```{r multiperiod, fig.width = 6, fig.height = 4}
df3 <- data.frame(
  x     = rep(1:3, each = 4),
  y     = c(1,2,3,4, 3,1,4,2, 2,4,1,3),
  group = rep(LETTERS[1:4], 3)
)

lbl_l <- df3[df3$x == 1, ]; lbl_r <- df3[df3$x == 3, ]

ggplot(df3, aes(x, y, group = group, fill = after_stat(avg_y))) +
  geom_bump_ribbon(alpha = 0.7) +
  scale_fill_viridis_c(guide = "none") +
  scale_y_reverse() +
  scale_x_continuous(limits = c(0.3, 3.7)) +
  geom_text(data = lbl_l, aes(x = 0.92, y = y, label = paste(y, group)),
            inherit.aes = FALSE, hjust = 1, size = 4) +
  geom_text(data = lbl_r, aes(x = 3.08, y = y, label = paste(group, y)),
            inherit.aes = FALSE, hjust = 0, size = 4) +
  theme_void()
```

## `geom_bump_line()`

```{r bumpline, fig.width = 6, fig.height = 4}
df <- data.frame(
  x     = rep(1:2, each = 5),
  y     = c(1, 2, 3, 4, 5, 3, 1, 5, 2, 4),
  group = rep(LETTERS[1:5], 2)
)

lbl_l <- df[df$x == 1, ]; lbl_r <- df[df$x == 2, ]

ggplot(df, aes(x, y, group = group, colour = after_stat(avg_y))) +
  geom_bump_line(linewidth = 1.2) +
  scale_colour_viridis_c(guide = "none") +
  scale_y_reverse() +
  scale_x_continuous(limits = c(0.3, 2.7)) +
  geom_text(data = lbl_l, aes(x = 0.92, y = y, label = paste(y, group)),
            inherit.aes = FALSE, hjust = 1, size = 4) +
  geom_text(data = lbl_r, aes(x = 2.08, y = y, label = paste(group, y)),
            inherit.aes = FALSE, hjust = 0, size = 4) +
  theme_void()
```

## mtcars

```{r mtcars, fig.width = 6, fig.height = 5}
mt <- mtcars[1:10, ]
mt$car <- rownames(mt)

mt_long <- data.frame(
  x     = rep(1:2, each = 10),
  y     = c(rank(-mt$mpg), rank(-mt$hp)),
  group = rep(mt$car, 2)
)

lbl_l <- mt_long[mt_long$x == 1, ]; lbl_r <- mt_long[mt_long$x == 2, ]

ggplot(mt_long, aes(x, y, group = group, fill = after_stat(avg_y))) +
  geom_bump_ribbon() +
  scale_fill_rank(limits = c(1, 10)) +
  scale_y_reverse() +
  scale_x_continuous(limits = c(0.1, 2.9)) +
  geom_text(data = lbl_l, aes(x = 0.93, y = y, label = paste(y, group)),
            inherit.aes = FALSE, hjust = 1, colour = "white", size = 2.8) +
  geom_text(data = lbl_r, aes(x = 2.07, y = y, label = paste(group, y)),
            inherit.aes = FALSE, hjust = 0, colour = "white", size = 2.8) +
  annotate("text", x = 1, y = -0.3, label = "MPG Rank", colour = "white",
           size = 4, fontface = "bold") +
  annotate("text", x = 2, y = -0.3, label = "HP Rank", colour = "white",
           size = 4, fontface = "bold") +
  labs(title = "mtcars: MPG rank vs HP rank") +
  theme_bump()
```

## Gallery

### 1. The Grammy Bump — slope chart

```{r grammy, fig.width = 7, fig.height = 6}
grammy <- data.frame(stringsAsFactors = FALSE,
  group     = c("Adele","Taylor Swift","Billie Eilish","Beyonce",
                "Daft Punk","Bruno Mars","Mumford & Sons","Beck"),
  rank_from = c(5, 3, 7, 2, 8, 4, 6, 1),
  rank_to   = c(1, 2, 3, 5, 6, 4, 8, 7))

df <- data.frame(
  x = rep(1:2, each = 8), y = c(grammy$rank_from, grammy$rank_to),
  group = rep(grammy$group, 2))
lbl_l <- df[df$x == 1, ]; lbl_r <- df[df$x == 2, ]

ggplot(df, aes(x, y, group = group, colour = after_stat(avg_y))) +
  geom_bump_line(linewidth = 1.5, smooth = 6) +
  geom_point(aes(x, y), data = df, inherit.aes = FALSE, size = 3, colour = "white") +
  scale_colour_gradientn(colours = c("#fbbf24","#f97316","#ec4899","#8b5cf6"),
                         guide = "none") +
  scale_y_reverse(breaks = 1:8) +
  scale_x_continuous(limits = c(0.2, 2.8), breaks = 1:2,
                     labels = c("Pre-Grammy\nSales Rank", "Post-Grammy\nSales Rank")) +
  geom_text(data = lbl_l, aes(x = 0.94, y = y, label = y), inherit.aes = FALSE,
            hjust = 1, colour = "grey90", size = 2.5) +
  geom_text(data = lbl_l, aes(x = 0.88, y = y, label = group), inherit.aes = FALSE,
            hjust = 1, colour = "grey90", size = 2.8) +
  geom_text(data = lbl_r, aes(x = 2.06, y = y, label = y), inherit.aes = FALSE,
            hjust = 0, colour = "grey90", size = 2.5) +
  geom_text(data = lbl_r, aes(x = 2.12, y = y, label = group), inherit.aes = FALSE,
            hjust = 0, colour = "grey90", size = 2.8) +
  labs(title = "THE GRAMMY BUMP",
       subtitle = "Album sales rank shift after winning Album of the Year") +
  theme_void() +
  theme(plot.background = element_rect(fill = "#111111", colour = NA),
        panel.background = element_rect(fill = "#111111", colour = NA),
        axis.text.x = element_text(colour = "#fbbf24", size = 10, face = "bold"),
        plot.title = element_text(colour = "#fbbf24", size = 18, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(colour = "grey60", size = 9, hjust = 0.5))
```

### 2. Programming Languages — multi-period lines

```{r languages, fig.width = 8, fig.height = 5}
langs <- c("Python","JavaScript","Java","TypeScript","Go","C++","Rust","PHP")
df <- data.frame(stringsAsFactors = FALSE,
  group = rep(langs, 5),
  x = rep(c(2016, 2018, 2020, 2022, 2024), each = 8),
  y = c(3,1,2,8,5,4,7,6, 2,1,3,6,4,5,7,8, 1,2,3,5,4,6,7,8,
        1,2,3,4,5,6,7,8, 1,3,4,2,5,6,7,8))
lbl_l <- df[df$x == 2016, ]; lbl_r <- df[df$x == 2024, ]

ggplot(df, aes(x, y, group = group, colour = group)) +
  geom_bump_line(linewidth = 1, smooth = 5) +
  geom_point(data = df[df$x %in% c(2016, 2024), ],
             aes(x, y, colour = group), inherit.aes = FALSE, size = 2.5) +
  scale_colour_manual(values = c(Python="#3776AB",JavaScript="#F7DF1E",Java="#ED8B00",
                                 TypeScript="#3178C6",Go="#00ADD8",`C++`="#00599C",
                                 Rust="#CE422B",PHP="#777BB4"), guide = "none") +
  scale_y_reverse(breaks = 1:8) +
  scale_x_continuous(limits = c(2013, 2027), breaks = c(2016,2018,2020,2022,2024)) +
  geom_text(data = lbl_l, aes(x = 2015.7, y = y, label = group),
            inherit.aes = FALSE, hjust = 1, size = 2.8, colour = "grey20") +
  geom_text(data = lbl_r, aes(x = 2024.3, y = y, label = group),
            inherit.aes = FALSE, hjust = 0, size = 2.8, colour = "grey20") +
  labs(title = "Programming Language Popularity on GitHub",
       subtitle = "Simulated rank by pull requests 2016–2024", x = NULL, y = "Rank") +
  theme_light(base_size = 10) +
  theme(panel.grid.minor = element_blank(), plot.title = element_text(face = "bold"))
```

### 3. Leading Causes of Death — multi-period ribbon

```{r health, fig.width = 9, fig.height = 5.5}
causes <- c("Heart Disease","Cancer","Accidents","Stroke",
            "Diabetes","Alzheimer's","Influenza","Kidney Disease")
df <- data.frame(stringsAsFactors = FALSE,
  group = rep(causes, 5),
  x = rep(c(2000, 2005, 2010, 2015, 2020), each = 8),
  y = c(1,2,5,3,6,7,4,8, 1,2,4,3,6,5,7,8, 1,2,3,4,7,5,6,8,
        1,2,3,5,7,4,6,8, 1,2,3,5,8,4,6,7))
lbl_l <- df[df$x == 2000, ]; lbl_r <- df[df$x == 2020, ]

ggplot(df, aes(x, y, group = group, fill = after_stat(avg_y))) +
  geom_bump_ribbon(alpha = 0.75, width = 0.65, smooth = 5) +
  scale_fill_gradientn(colours = c("#991b1b","#dc2626","#f59e0b","#84cc16","#0ea5e9","#6366f1"),
                       guide = "none") +
  scale_y_reverse(breaks = 1:8) +
  scale_x_continuous(limits = c(1995.5, 2024.5), breaks = c(2000,2005,2010,2015,2020)) +
  geom_text(data = lbl_l, aes(x = 1999.5, y = y, label = group),
            inherit.aes = FALSE, hjust = 1, size = 2.5, colour = "grey30") +
  geom_text(data = lbl_l, aes(x = 1999.8, y = y, label = y),
            inherit.aes = FALSE, hjust = 1, size = 2.2, colour = "grey50") +
  geom_text(data = lbl_r, aes(x = 2020.5, y = y, label = group),
            inherit.aes = FALSE, hjust = 0, size = 2.5, colour = "grey30") +
  geom_text(data = lbl_r, aes(x = 2020.2, y = y, label = y),
            inherit.aes = FALSE, hjust = 0, size = 2.2, colour = "grey50") +
  labs(title = "Leading Causes of Death in the U.S.",
       subtitle = "Simulated rank changes 2000–2020", x = NULL, y = "Rank") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold"))
```

### 4. Quality of Life — slope ribbon with entries/exits

```{r quality, fig.width = 6, fig.height = 7}
both <- data.frame(stringsAsFactors = FALSE,
  group     = c("Netherlands","Denmark","Switzerland","Norway","Finland",
                "Germany","Austria","Australia"),
  rank_from = c(12, 1, 2, 5, 6, 7, 4, 8),
  rank_to   = c(1, 2, 3, 5, 6, 7, 8, 10))
exit_only  <- data.frame(group = c("Sweden","Canada"), rank_from = c(3, 9))
enter_only <- data.frame(group = c("New Zealand","Japan"), rank_to = c(4, 9))

df <- data.frame(x = rep(1:2, each = nrow(both)),
                 y = c(both$rank_from, both$rank_to),
                 group = rep(both$group, 2))
lbl_l <- df[df$x == 1, ]; lbl_r <- df[df$x == 2, ]

ggplot(df, aes(x, y, group = group, fill = after_stat(avg_y))) +
  geom_bump_ribbon(alpha = 0.8, width = 0.55) +
  scale_fill_gradientn(colours = c("#059669","#34d399","#fbbf24","#f97316","#dc2626"),
                       guide = "none") +
  scale_y_reverse(breaks = 1:12) +
  scale_x_continuous(limits = c(0.1, 2.9), breaks = 1:2, labels = c("2015","2025")) +
  geom_text(data = lbl_l, aes(x = 0.94, y = y, label = y),
            inherit.aes = FALSE, hjust = 1, colour = "grey30", size = 3) +
  geom_text(data = lbl_l, aes(x = 0.88, y = y, label = group),
            inherit.aes = FALSE, hjust = 1, colour = "grey20", size = 3.2, fontface = "bold") +
  geom_text(data = lbl_r, aes(x = 2.06, y = y, label = y),
            inherit.aes = FALSE, hjust = 0, colour = "grey30", size = 3) +
  geom_text(data = lbl_r, aes(x = 2.12, y = y, label = group),
            inherit.aes = FALSE, hjust = 0, colour = "grey20", size = 3.2, fontface = "bold") +
  geom_text(data = exit_only, aes(x = 0.94, y = rank_from, label = rank_from),
            inherit.aes = FALSE, hjust = 1, colour = "grey65", size = 3) +
  geom_text(data = exit_only, aes(x = 0.88, y = rank_from, label = group),
            inherit.aes = FALSE, hjust = 1, colour = "grey65", size = 3.2) +
  geom_text(data = enter_only, aes(x = 2.06, y = rank_to, label = rank_to),
            inherit.aes = FALSE, hjust = 0, colour = "grey65", size = 3) +
  geom_text(data = enter_only, aes(x = 2.12, y = rank_to, label = group),
            inherit.aes = FALSE, hjust = 0, colour = "grey65", size = 3.2) +
  annotate("text", x = 1, y = -0.5, label = "2015", colour = "grey30",
           size = 5.5, fontface = "bold") +
  annotate("text", x = 2, y = -0.5, label = "2025", colour = "grey30",
           size = 5.5, fontface = "bold") +
  labs(title = "Top Countries by Quality of Life",
       subtitle = "Netherlands climbs 11 places",
       caption = "Grey = entered/exited top 10") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
        plot.subtitle = element_text(colour = "grey50", size = 9, hjust = 0.5),
        plot.caption = element_text(colour = "grey60", size = 7, hjust = 1))
```

## Parameters

| Parameter | Default | Description |
|-----------|---------|-------------|
| `smooth` | `8` | Steepness of the sigmoid curve. Higher = sharper S-shape |
| `width` | `0.8` | Ribbon full width in data units (`geom_bump_ribbon()` only) |
| `n` | `100` | Interpolation points per segment |
| `alpha` | `0.85` | Ribbon transparency |

## Computed variables

| Variable | Description |
|----------|-------------|
| `avg_y` | Mean of y values in the group, inverse-transformed to original data space. Works with `scale_y_reverse()`. |
| `ymin` | Lower ribbon boundary (`geom_bump_ribbon()` only) |
| `ymax` | Upper ribbon boundary (`geom_bump_ribbon()` only) |

## Convenience functions

| Function | Description |
|----------|-------------|
| `scale_fill_rank()` | Green-yellow-red gradient scale with `guide = "none"` default |
| `theme_bump()` | Dark background theme for rank comparison infographics |

## Architecture

```
User data (x, y, group)
  → StatBumpRibbon$compute_group() / StatBumpLine$compute_group()
    → sigmoid_path() for upper/lower edges (ribbon) or centerline (line)
    → returns data.frame(x, y, ymin, ymax, avg_y)
  → GeomRibbon / GeomPath renders
```

## Dependencies

**Hard:** ggplot2 (>= 3.5.0), rlang, scales — rlang and scales are already ggplot2 dependencies.

## License

MIT
