---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dev = "png",
  dev.args = list(type = "cairo"),
  dpi = 300
)
```

# ggbumpribbon

<!-- badges: start -->
[![R-CMD-check](https://github.com/sondreskarsten/ggbumpribbon/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sondreskarsten/ggbumpribbon/actions/workflows/R-CMD-check.yaml)
[![r-universe](https://sondreskarsten.r-universe.dev/badges/ggbumpribbon)](https://sondreskarsten.r-universe.dev/ggbumpribbon)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Sigmoid-curved filled ribbons and lines for rank comparison charts in ggplot2. Two geoms — `geom_bump_ribbon()` for filled areas and `geom_bump_line()` for stroked paths — with C1-continuous segment joins via logistic sigmoid or cubic Hermite interpolation.

```{r reputation, echo = FALSE, fig.width = 7, fig.height = 17}
library(ggplot2)
library(ggbumpribbon)
library(ggflags)
library(countrycode)

ranks <- data.frame(stringsAsFactors = FALSE,
  country   = c("Switzerland","Norway","Sweden","Canada","Denmark","New Zealand","Finland",
                "Australia","Ireland","Netherlands","Austria","Japan","Spain","Italy","Belgium",
                "Portugal","Greece","UK","Singapore","France","Germany","Czechia","Thailand",
                "Poland","South Korea","Malaysia","Indonesia","Peru","Brazil","U.S.","Ukraine",
                "Philippines","Morocco","Chile","Hungary","Argentina","Vietnam","Egypt","UAE",
                "South Africa","Mexico","Romania","India","Turkey","Qatar","Algeria","Ethiopia",
                "Colombia","Kazakhstan","Nigeria","Bangladesh","Israel","Saudi Arabia","Pakistan",
                "China","Iran","Iraq","Russia"),
  rank_from = c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24,25,26,27,28,
                29,30,31,32,33,34,35,36,37,38,39,40,41,42,43,44,45,51,47,49,50,52,53,54,55,56,
                57,58,59,60),
  rank_to   = c(1,3,4,2,6,7,5,11,10,9,12,8,14,13,17,15,16,18,19,21,20,25,24,23,31,29,34,27,
                28,48,26,33,30,35,32,38,37,36,40,42,39,41,45,43,44,46,51,50,49,52,54,55,53,56,
                57,59,58,60))

exit_only  <- data.frame(country = c("Cuba","Venezuela"),  rank_from = c(46,48), stringsAsFactors = FALSE)
enter_only <- data.frame(country = c("Taiwan","Kuwait"),   rank_to   = c(22,47), stringsAsFactors = FALSE)

ov <- c("U.S."="us","UK"="gb","South Korea"="kr","Czechia"="cz","Taiwan"="tw","UAE"="ae")
iso <- function(x) ifelse(x %in% names(ov), ov[x],
  tolower(countrycode(x, "country.name", "iso2c", warn = FALSE)))

ranks$iso2      <- iso(ranks$country)
exit_only$iso2  <- iso(exit_only$country)
enter_only$iso2 <- iso(enter_only$country)

ranks_long <- data.frame(
  x       = rep(1:2, each = nrow(ranks)),
  y       = c(ranks$rank_from, ranks$rank_to),
  group   = rep(ranks$country, 2),
  country = rep(ranks$country, 2),
  iso2    = rep(ranks$iso2, 2))

lbl_l <- ranks_long[ranks_long$x == 1, ]
lbl_r <- ranks_long[ranks_long$x == 2, ]

ggplot(ranks_long, aes(x, y, group = group, fill = after_stat(avg_y))) +
  geom_bump_ribbon(alpha = 0.85, width = 0.8) +
  scale_fill_gradientn(
    colours = c("#2ecc71","#a8e063","#f7dc6f","#f0932b","#eb4d4b","#c0392b"),
    guide = "none") +
  scale_y_reverse(expand = expansion(mult = c(0.015, 0.015))) +
  scale_x_continuous(limits = c(0.15, 2.85)) +
  geom_text(data = lbl_l, aes(x = 0.94, y = y, label = y),
            inherit.aes = FALSE, hjust = 1, colour = "white", size = 2.2) +
  geom_flag(data = lbl_l, aes(x = 0.88, y = y, country = iso2),
            inherit.aes = FALSE, size = 3) +
  geom_text(data = lbl_l, aes(x = 0.82, y = y, label = country),
            inherit.aes = FALSE, hjust = 1, colour = "white", size = 2.2) +
  geom_text(data = lbl_r, aes(x = 2.06, y = y, label = y),
            inherit.aes = FALSE, hjust = 0, colour = "white", size = 2.2) +
  geom_flag(data = lbl_r, aes(x = 2.12, y = y, country = iso2),
            inherit.aes = FALSE, size = 3) +
  geom_text(data = lbl_r, aes(x = 2.18, y = y, label = country),
            inherit.aes = FALSE, hjust = 0, colour = "white", size = 2.2) +
  geom_text(data = exit_only, aes(x = 0.94, y = rank_from, label = rank_from),
            inherit.aes = FALSE, hjust = 1, colour = "grey55", size = 2.2) +
  geom_flag(data = exit_only, aes(x = 0.88, y = rank_from, country = iso2),
            inherit.aes = FALSE, size = 3) +
  geom_text(data = exit_only, aes(x = 0.82, y = rank_from, label = country),
            inherit.aes = FALSE, hjust = 1, colour = "grey55", size = 2.2) +
  geom_text(data = enter_only, aes(x = 2.06, y = rank_to, label = rank_to),
            inherit.aes = FALSE, hjust = 0, colour = "grey55", size = 2.2) +
  geom_flag(data = enter_only, aes(x = 2.12, y = rank_to, country = iso2),
            inherit.aes = FALSE, size = 3) +
  geom_text(data = enter_only, aes(x = 2.18, y = rank_to, label = country),
            inherit.aes = FALSE, hjust = 0, colour = "grey55", size = 2.2) +
  annotate("text", x = 1, y = -1.5, label = "2024 Rank",
           colour = "white", size = 4.5, fontface = "bold") +
  annotate("text", x = 2, y = -1.5, label = "2025 Rank",
           colour = "white", size = 4.5, fontface = "bold") +
  labs(title    = "COUNTRIES WITH THE BEST REPUTATIONS IN 2025",
       subtitle = "Reputation Lab ranked the reputations of 60 leading economies\nin 2025, shedding light on their international standing.",
       caption  = "Source: Reputation Lab | Made with ggbumpribbon") +
  theme_bump()
```

<details>
<summary>Code to reproduce (requires ggflags + countrycode)</summary>

```{r ref.label = "reputation", eval = FALSE}
```

Flags require [ggflags](https://github.com/jimjam-slam/ggflags):
`install.packages("ggflags", repos = c("https://jimjam-slam.r-universe.dev", "https://cloud.r-project.org"))`

</details>

```{r gdp, echo = FALSE, fig.width = 7, fig.height = 10}
library(ggplot2)
library(ggbumpribbon)
library(ggflags)
library(countrycode)

both <- data.frame(stringsAsFactors = FALSE,
  country   = c("U.S.","Japan","Germany","France","UK","Italy","China","Canada",
                "Mexico","Spain","Netherlands","India","Saudi Arabia","Australia","Brazil"),
  rank_from = c(1,2,3,4,5,6,7,8,9,11,12,13,14,15,16),
  rank_to   = c(1,4,3,7,6,8,2,10,13,12,18,5,19,15,11),
  gdp_1980  = c(2.9,1.1,0.857,0.695,0.605,0.480,0.304,0.276,
                0.242,0.231,0.194,0.186,0.165,0.163,0.146),
  gdp_2025  = c(30.6,4.3,5.0,3.4,4.0,2.5,19.4,2.3,1.9,1.9,1.3,4.1,1.3,1.8,2.3))

exit_only_gdp <- data.frame(stringsAsFactors = FALSE,
  country   = c("Argentina","Sweden","Belgium","Switzerland","Iran"),
  rank_from = c(10,17,18,19,20),
  gdp_1980  = c(0.234,0.140,0.123,0.122,0.117))

enter_only_gdp <- data.frame(stringsAsFactors = FALSE,
  country   = c("Russia","S. Korea","Turkey","Indonesia","Poland"),
  rank_to   = c(9,14,16,17,20),
  gdp_2025  = c(2.5,1.9,1.6,1.4,1.0))

ov2 <- c("U.S."="us","UK"="gb","S. Korea"="kr","Turkey"="tr")
iso2 <- function(x) ifelse(x %in% names(ov2), ov2[x],
  tolower(countrycode(x, "country.name", "iso2c", warn = FALSE)))

both$iso2           <- iso2(both$country)
exit_only_gdp$iso2  <- iso2(exit_only_gdp$country)
enter_only_gdp$iso2 <- iso2(enter_only_gdp$country)

both_long <- data.frame(
  x       = rep(c(1, 1.35, 1.65, 2), each = nrow(both)),
  y       = c(both$rank_from, both$rank_from, both$rank_to, both$rank_to),
  group   = rep(both$country, 4),
  country = rep(both$country, 4),
  iso2    = rep(both$iso2, 4))

lbl_l_gdp <- both_long[both_long$x == 1, ]
lbl_r_gdp <- both_long[both_long$x == 2, ]

fmt_gdp <- function(x) ifelse(x >= 1,
  paste0("$", formatC(x, format = "f", digits = 1), "T"),
  paste0("$", round(x * 1000), "B"))

both$gdp_l_label          <- fmt_gdp(both$gdp_1980)
both$gdp_r_label          <- fmt_gdp(both$gdp_2025)
exit_only_gdp$gdp_label   <- fmt_gdp(exit_only_gdp$gdp_1980)
enter_only_gdp$gdp_label  <- fmt_gdp(enter_only_gdp$gdp_2025)

gdp_max   <- 31
bar_left  <- 2.22
bar_width <- 0.55
bars_r <- data.frame(y = both$rank_to, xmin = bar_left,
  xmax = bar_left + bar_width * (both$gdp_2025 / gdp_max), gdp = both$gdp_2025)
bars_enter <- data.frame(y = enter_only_gdp$rank_to, xmin = bar_left,
  xmax = bar_left + bar_width * (enter_only_gdp$gdp_2025 / gdp_max), gdp = enter_only_gdp$gdp_2025)

row_bg_l <- data.frame(y = 1:20, xmin = 0.15, xmax = 0.98,
  fill = ifelse(1:20 %% 2 == 0, "#0a1a3a", "#0e2248"))
row_bg_r <- data.frame(y = 1:20, xmin = 2.02, xmax = 2.85,
  fill = ifelse(1:20 %% 2 == 0, "#0a1a3a", "#0e2248"))

bg <- "#0b1a38"

ggplot() +
  geom_rect(data = row_bg_l,
    aes(xmin = xmin, xmax = xmax, ymin = y - 0.48, ymax = y + 0.48),
    fill = row_bg_l$fill, colour = NA) +
  geom_rect(data = row_bg_r,
    aes(xmin = xmin, xmax = xmax, ymin = y - 0.48, ymax = y + 0.48),
    fill = row_bg_r$fill, colour = NA) +
  geom_rect(data = bars_r,
    aes(xmin = xmin, xmax = xmax, ymin = y - 0.35, ymax = y + 0.35),
    fill = scales::seq_gradient_pal("#3b82f6", "#ec4899")(bars_r$gdp / gdp_max),
    colour = NA, alpha = 0.7) +
  geom_rect(data = bars_enter,
    aes(xmin = xmin, xmax = xmax, ymin = y - 0.35, ymax = y + 0.35),
    fill = scales::seq_gradient_pal("#3b82f6", "#ec4899")(bars_enter$gdp / gdp_max),
    colour = NA, alpha = 0.5) +
  geom_bump_line(data = both_long,
    aes(x = x, y = y, group = group, colour = after_stat(avg_y)),
    linewidth = 0.7, smooth = 10) +
  scale_colour_gradientn(
    colours = c("#fbbf24","#f59e0b","#22d3ee","#818cf8","#a78bfa"),
    guide = "none") +
  scale_y_reverse(expand = expansion(mult = c(0.03, 0.03))) +
  scale_x_continuous(limits = c(0.12, 2.88)) +
  geom_text(data = lbl_l_gdp, aes(x = 0.20, y = y, label = y),
    inherit.aes = FALSE, hjust = 0, colour = "white", size = 3.2, fontface = "bold") +
  geom_flag(data = lbl_l_gdp, aes(x = 0.30, y = y, country = iso2),
    inherit.aes = FALSE, size = 2.5) +
  geom_text(data = lbl_l_gdp, aes(x = 0.38, y = y, label = country),
    inherit.aes = FALSE, hjust = 0, colour = "white", size = 2.8, fontface = "bold") +
  geom_text(data = both, aes(x = 0.95, y = rank_from, label = gdp_l_label),
    inherit.aes = FALSE, hjust = 1, colour = "grey75", size = 2.5) +
  geom_flag(data = lbl_r_gdp, aes(x = 2.07, y = y, country = iso2),
    inherit.aes = FALSE, size = 2.5) +
  geom_text(data = lbl_r_gdp, aes(x = 2.15, y = y, label = country),
    inherit.aes = FALSE, hjust = 0, colour = "white", size = 2.8, fontface = "bold") +
  geom_text(data = both, aes(x = 2.78, y = rank_to, label = gdp_r_label),
    inherit.aes = FALSE, hjust = 1, colour = "white", size = 2.5) +
  geom_text(data = lbl_r_gdp, aes(x = 2.82, y = y, label = y),
    inherit.aes = FALSE, hjust = 0, colour = "white", size = 3.2, fontface = "bold") +
  geom_text(data = exit_only_gdp, aes(x = 0.20, y = rank_from, label = rank_from),
    inherit.aes = FALSE, hjust = 0, colour = "grey50", size = 3.2, fontface = "bold") +
  geom_flag(data = exit_only_gdp, aes(x = 0.30, y = rank_from, country = iso2),
    inherit.aes = FALSE, size = 2.5) +
  geom_text(data = exit_only_gdp, aes(x = 0.38, y = rank_from, label = country),
    inherit.aes = FALSE, hjust = 0, colour = "grey50", size = 2.8, fontface = "bold") +
  geom_text(data = exit_only_gdp, aes(x = 0.95, y = rank_from, label = gdp_label),
    inherit.aes = FALSE, hjust = 1, colour = "grey45", size = 2.5) +
  geom_flag(data = enter_only_gdp, aes(x = 2.07, y = rank_to, country = iso2),
    inherit.aes = FALSE, size = 2.5) +
  geom_text(data = enter_only_gdp, aes(x = 2.15, y = rank_to, label = country),
    inherit.aes = FALSE, hjust = 0, colour = "grey50", size = 2.8, fontface = "bold") +
  geom_text(data = enter_only_gdp, aes(x = 2.78, y = rank_to, label = gdp_label),
    inherit.aes = FALSE, hjust = 1, colour = "grey50", size = 2.5) +
  geom_text(data = enter_only_gdp, aes(x = 2.82, y = rank_to, label = rank_to),
    inherit.aes = FALSE, hjust = 0, colour = "grey50", size = 3.2, fontface = "bold") +
  annotate("text", x = 0.55, y = -0.8, label = "1980",
    colour = "#60a5fa", size = 7, fontface = "bold") +
  annotate("text", x = 2.45, y = -0.8, label = "2025",
    colour = "#60a5fa", size = 7, fontface = "bold") +
  annotate("segment", x = 0.85, xend = 2.15, y = -0.8, yend = -0.8,
    colour = "grey40", linewidth = 0.3) +
  labs(title   = "TOP 20 ECONOMIES",
       caption = "Source: IMF, World Economic Outlook October 2025 | Made with ggbumpribbon") +
  theme_void() +
  theme(plot.background  = element_rect(fill = bg, colour = NA),
        panel.background = element_rect(fill = bg, colour = NA),
        plot.title       = element_text(colour = "#93c5fd", size = 22, face = "bold",
                                        hjust = 0.5, margin = margin(t = 12, b = 2)),
        plot.caption     = element_text(colour = "grey45", size = 6,
                                        hjust = 0.5, margin = margin(t = 8, b = 5)),
        plot.margin      = margin(8, 8, 8, 8))
```

<details>
<summary>Code to reproduce (requires ggflags + countrycode)</summary>

```{r ref.label = "gdp", eval = FALSE}
```

</details>

## Installation

```{r installation, eval = FALSE}
install.packages("ggbumpribbon",
  repos = c("https://sondreskarsten.r-universe.dev", "https://cloud.r-project.org"))
```

Or from GitHub:

```{r installation2, eval = FALSE}
# install.packages("pak")
pak::pak("sondreskarsten/ggbumpribbon")
```

## Usage

### Minimal example

```{r basic, fig.width = 6, fig.height = 4}
library(ggplot2)
library(ggbumpribbon)

df <- data.frame(
  x     = rep(1:2, each = 5),
  y     = c(1, 2, 3, 4, 5, 3, 1, 5, 2, 4),
  group = rep(LETTERS[1:5], 2)
)

lbl_l <- df[df$x == 1, ]; lbl_r <- df[df$x == 2, ]

ggplot(df, aes(x, y, group = group, fill = after_stat(avg_y))) +
  geom_bump_ribbon(alpha = 0.85) +
  scale_fill_gradientn(colours = c("#2ecc71", "#f7dc6f", "#eb4d4b"), guide = "none") +
  scale_y_reverse() +
  scale_x_continuous(limits = c(0.3, 2.7)) +
  geom_text(data = lbl_l, aes(x = 0.92, y = y, label = paste(y, group)),
            inherit.aes = FALSE, hjust = 1, size = 4) +
  geom_text(data = lbl_r, aes(x = 2.08, y = y, label = paste(group, y)),
            inherit.aes = FALSE, hjust = 0, size = 4) +
  theme_void()
```

### Lines

The line counterpart to `geom_bump_ribbon()`. Uses sigmoid curves between rank positions but renders as stroked paths via `GeomPath` instead of filled areas. Map `colour = after_stat(avg_y)` instead of `fill`.

```{r bumpline, fig.width = 6, fig.height = 4}
library(ggplot2)
library(ggbumpribbon)

df <- data.frame(
  x     = rep(1:2, each = 5),
  y     = c(1, 2, 3, 4, 5, 3, 1, 5, 2, 4),
  group = rep(LETTERS[1:5], 2)
)

lbl_l <- df[df$x == 1, ]; lbl_r <- df[df$x == 2, ]

ggplot(df, aes(x, y, group = group, colour = after_stat(avg_y))) +
  geom_bump_line(linewidth = 1.2) +
  scale_colour_viridis_c(guide = "none") +
  scale_y_reverse() +
  scale_x_continuous(limits = c(0.3, 2.7)) +
  geom_text(data = lbl_l, aes(x = 0.92, y = y, label = paste(y, group)),
            inherit.aes = FALSE, hjust = 1, size = 4) +
  geom_text(data = lbl_r, aes(x = 2.08, y = y, label = paste(group, y)),
            inherit.aes = FALSE, hjust = 0, size = 4) +
  theme_void()
```

### Multi-period

Ribbons and lines chain automatically across 3+ time points:

```{r multiperiod, fig.width = 6, fig.height = 4}
library(ggplot2)
library(ggbumpribbon)

df3 <- data.frame(
  x     = rep(1:3, each = 4),
  y     = c(1,2,3,4, 3,1,4,2, 2,4,1,3),
  group = rep(LETTERS[1:4], 3)
)

lbl_l <- df3[df3$x == 1, ]; lbl_r <- df3[df3$x == 3, ]

ggplot(df3, aes(x, y, group = group, fill = after_stat(avg_y))) +
  geom_bump_ribbon(alpha = 0.7) +
  scale_fill_viridis_c(guide = "none") +
  scale_y_reverse() +
  scale_x_continuous(limits = c(0.3, 3.7)) +
  geom_text(data = lbl_l, aes(x = 0.92, y = y, label = paste(y, group)),
            inherit.aes = FALSE, hjust = 1, size = 4) +
  geom_text(data = lbl_r, aes(x = 3.08, y = y, label = paste(group, y)),
            inherit.aes = FALSE, hjust = 0, size = 4) +
  theme_void()
```

### Multi-bend curves

The number of bends is controlled entirely by the data shape, not by parameters. Two x-values per group produce one sigmoid. Four x-values — with the middle two holding position at the departure and arrival y — produce the "exit-channel-enter" pattern seen in the GDP hero above:

```r
# 1 sigmoid:  x = c(1, 2)
# 3 sigmoids: x = c(1, 1.35, 1.65, 2),  y = c(from, from, to, to)
```

Adjusting the gap between the middle x-values (`1.3`/`1.7` vs `1.45`/`1.55`) controls how narrow the central channel is.

### Real data (mtcars)

```{r mtcars, fig.width = 6, fig.height = 5}
library(ggplot2)
library(ggbumpribbon)

mt <- mtcars[1:10, ]
mt$car <- rownames(mt)

mt_long <- data.frame(
  x     = rep(1:2, each = 10),
  y     = c(rank(-mt$mpg, ties.method = "first"), rank(-mt$hp, ties.method = "first")),
  group = rep(mt$car, 2)
)

lbl_l <- mt_long[mt_long$x == 1, ]; lbl_r <- mt_long[mt_long$x == 2, ]

ggplot(mt_long, aes(x, y, group = group, fill = after_stat(avg_y))) +
  geom_bump_ribbon() +
  scale_fill_rank(limits = c(1, 10)) +
  scale_y_reverse() +
  scale_x_continuous(limits = c(0.1, 2.9)) +
  geom_text(data = lbl_l, aes(x = 0.93, y = y, label = paste(y, group)),
            inherit.aes = FALSE, hjust = 1, colour = "white", size = 2.8) +
  geom_text(data = lbl_r, aes(x = 2.07, y = y, label = paste(group, y)),
            inherit.aes = FALSE, hjust = 0, colour = "white", size = 2.8) +
  annotate("text", x = 1, y = -0.3, label = "MPG Rank", colour = "white",
           size = 4, fontface = "bold") +
  annotate("text", x = 2, y = -0.3, label = "HP Rank", colour = "white",
           size = 4, fontface = "bold") +
  labs(title = "mtcars: MPG rank vs HP rank") +
  theme_bump()
```

## Gallery

### 1. The Grammy Bump — slope chart

```{r grammy, fig.width = 7, fig.height = 6}
library(ggplot2)
library(ggbumpribbon)

grammy <- data.frame(stringsAsFactors = FALSE,
  group     = c("Adele","Taylor Swift","Billie Eilish","Beyonce",
                "Daft Punk","Bruno Mars","Mumford & Sons","Beck"),
  rank_from = c(5, 3, 7, 2, 8, 4, 6, 1),
  rank_to   = c(1, 2, 3, 5, 6, 4, 8, 7))

df <- data.frame(
  x = rep(1:2, each = 8), y = c(grammy$rank_from, grammy$rank_to),
  group = rep(grammy$group, 2))
lbl_l <- df[df$x == 1, ]; lbl_r <- df[df$x == 2, ]

ggplot(df, aes(x, y, group = group, colour = after_stat(avg_y))) +
  geom_bump_line(linewidth = 1.5, smooth = 6) +
  geom_point(aes(x, y), data = df, inherit.aes = FALSE, size = 3, colour = "white") +
  scale_colour_gradientn(colours = c("#fbbf24","#f97316","#ec4899","#8b5cf6"),
                         guide = "none") +
  scale_y_reverse(breaks = 1:8) +
  scale_x_continuous(limits = c(0.2, 2.8), breaks = 1:2,
                     labels = c("Pre-Grammy\nSales Rank", "Post-Grammy\nSales Rank")) +
  geom_text(data = lbl_l, aes(x = 0.94, y = y, label = y), inherit.aes = FALSE,
            hjust = 1, colour = "grey90", size = 2.5) +
  geom_text(data = lbl_l, aes(x = 0.88, y = y, label = group), inherit.aes = FALSE,
            hjust = 1, colour = "grey90", size = 2.8) +
  geom_text(data = lbl_r, aes(x = 2.06, y = y, label = y), inherit.aes = FALSE,
            hjust = 0, colour = "grey90", size = 2.5) +
  geom_text(data = lbl_r, aes(x = 2.12, y = y, label = group), inherit.aes = FALSE,
            hjust = 0, colour = "grey90", size = 2.8) +
  labs(title = "THE GRAMMY BUMP",
       subtitle = "Album sales rank shift after winning Album of the Year") +
  theme_void() +
  theme(plot.background = element_rect(fill = "#111111", colour = NA),
        panel.background = element_rect(fill = "#111111", colour = NA),
        axis.text.x = element_text(colour = "#fbbf24", size = 10, face = "bold"),
        plot.title = element_text(colour = "#fbbf24", size = 18, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(colour = "grey60", size = 9, hjust = 0.5))
```

### 2. Programming Languages — multi-period lines

```{r languages, fig.width = 8, fig.height = 5}
library(ggplot2)
library(ggbumpribbon)

langs <- c("Python","JavaScript","Java","TypeScript","Go","C++","Rust","PHP")
df <- data.frame(stringsAsFactors = FALSE,
  group = rep(langs, 5),
  x = rep(c(2016, 2018, 2020, 2022, 2024), each = 8),
  y = c(3,1,2,8,5,4,7,6, 2,1,3,6,4,5,7,8, 1,2,3,5,4,6,7,8,
        1,2,3,4,5,6,7,8, 1,3,4,2,5,6,7,8))
lbl_l <- df[df$x == 2016, ]; lbl_r <- df[df$x == 2024, ]

ggplot(df, aes(x, y, group = group, colour = group)) +
  geom_bump_line(linewidth = 1, smooth = 5) +
  geom_point(data = df[df$x %in% c(2016, 2024), ],
             aes(x, y, colour = group), inherit.aes = FALSE, size = 2.5) +
  scale_colour_manual(values = c(Python="#3776AB",JavaScript="#F7DF1E",Java="#ED8B00",
                                 TypeScript="#3178C6",Go="#00ADD8",`C++`="#00599C",
                                 Rust="#CE422B",PHP="#777BB4"), guide = "none") +
  scale_y_reverse(breaks = 1:8) +
  scale_x_continuous(limits = c(2013, 2027), breaks = c(2016,2018,2020,2022,2024)) +
  geom_text(data = lbl_l, aes(x = 2015.7, y = y, label = group),
            inherit.aes = FALSE, hjust = 1, size = 2.8, colour = "grey20") +
  geom_text(data = lbl_r, aes(x = 2024.3, y = y, label = group),
            inherit.aes = FALSE, hjust = 0, size = 2.8, colour = "grey20") +
  labs(title = "Programming Language Popularity on GitHub",
       subtitle = "Simulated rank by pull requests 2016-2024", x = NULL, y = "Rank") +
  theme_light(base_size = 10) +
  theme(panel.grid.minor = element_blank(), plot.title = element_text(face = "bold"))
```

### 3. Leading Causes of Death — multi-period ribbon

```{r health, fig.width = 9, fig.height = 5.5}
library(ggplot2)
library(ggbumpribbon)

causes <- c("Heart Disease","Cancer","Accidents","Stroke",
            "Diabetes","Alzheimer's","Influenza","Kidney Disease")
df <- data.frame(stringsAsFactors = FALSE,
  group = rep(causes, 5),
  x = rep(c(2000, 2005, 2010, 2015, 2020), each = 8),
  y = c(1,2,5,3,6,7,4,8, 1,2,4,3,6,5,7,8, 1,2,3,4,7,5,6,8,
        1,2,3,5,7,4,6,8, 1,2,3,5,8,4,6,7))
lbl_l <- df[df$x == 2000, ]; lbl_r <- df[df$x == 2020, ]

ggplot(df, aes(x, y, group = group, fill = after_stat(avg_y))) +
  geom_bump_ribbon(alpha = 0.75, width = 0.65, smooth = 5) +
  scale_fill_gradientn(colours = c("#991b1b","#dc2626","#f59e0b","#84cc16","#0ea5e9","#6366f1"),
                       guide = "none") +
  scale_y_reverse(breaks = 1:8) +
  scale_x_continuous(limits = c(1995.5, 2024.5), breaks = c(2000,2005,2010,2015,2020)) +
  geom_text(data = lbl_l, aes(x = 1999.5, y = y, label = group),
            inherit.aes = FALSE, hjust = 1, size = 2.5, colour = "grey30") +
  geom_text(data = lbl_l, aes(x = 1999.8, y = y, label = y),
            inherit.aes = FALSE, hjust = 1, size = 2.2, colour = "grey50") +
  geom_text(data = lbl_r, aes(x = 2020.5, y = y, label = group),
            inherit.aes = FALSE, hjust = 0, size = 2.5, colour = "grey30") +
  geom_text(data = lbl_r, aes(x = 2020.2, y = y, label = y),
            inherit.aes = FALSE, hjust = 0, size = 2.2, colour = "grey50") +
  labs(title = "Leading Causes of Death in the U.S.",
       subtitle = "Simulated rank changes 2000-2020", x = NULL, y = "Rank") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold"))
```

### 4. Quality of Life — slope ribbon with entries/exits

```{r quality, fig.width = 6, fig.height = 7}
library(ggplot2)
library(ggbumpribbon)

both <- data.frame(stringsAsFactors = FALSE,
  group     = c("Netherlands","Denmark","Switzerland","Norway","Finland",
                "Germany","Austria","Australia"),
  rank_from = c(12, 1, 2, 5, 6, 7, 4, 8),
  rank_to   = c(1, 2, 3, 5, 6, 7, 8, 10))
exit_only  <- data.frame(group = c("Sweden","Canada"), rank_from = c(3, 9))
enter_only <- data.frame(group = c("New Zealand","Japan"), rank_to = c(4, 9))

df <- data.frame(x = rep(1:2, each = nrow(both)),
                 y = c(both$rank_from, both$rank_to),
                 group = rep(both$group, 2))
lbl_l <- df[df$x == 1, ]; lbl_r <- df[df$x == 2, ]

ggplot(df, aes(x, y, group = group, fill = after_stat(avg_y))) +
  geom_bump_ribbon(alpha = 0.8, width = 0.55) +
  scale_fill_gradientn(colours = c("#059669","#34d399","#fbbf24","#f97316","#dc2626"),
                       guide = "none") +
  scale_y_reverse(breaks = 1:12) +
  scale_x_continuous(limits = c(0.1, 2.9), breaks = 1:2, labels = c("2015","2025")) +
  geom_text(data = lbl_l, aes(x = 0.94, y = y, label = y),
            inherit.aes = FALSE, hjust = 1, colour = "grey30", size = 3) +
  geom_text(data = lbl_l, aes(x = 0.88, y = y, label = group),
            inherit.aes = FALSE, hjust = 1, colour = "grey20", size = 3.2, fontface = "bold") +
  geom_text(data = lbl_r, aes(x = 2.05, y = y, label = y),
            inherit.aes = FALSE, hjust = 0, colour = "grey30", size = 3) +
  geom_text(data = lbl_r, aes(x = 2.14, y = y, label = group),
            inherit.aes = FALSE, hjust = 0, colour = "grey20", size = 3.2, fontface = "bold") +
  geom_text(data = exit_only, aes(x = 0.94, y = rank_from, label = rank_from),
            inherit.aes = FALSE, hjust = 1, colour = "grey65", size = 3) +
  geom_text(data = exit_only, aes(x = 0.88, y = rank_from, label = group),
            inherit.aes = FALSE, hjust = 1, colour = "grey65", size = 3.2) +
  geom_text(data = enter_only, aes(x = 2.05, y = rank_to, label = rank_to),
            inherit.aes = FALSE, hjust = 0, colour = "grey65", size = 3) +
  geom_text(data = enter_only, aes(x = 2.14, y = rank_to, label = group),
            inherit.aes = FALSE, hjust = 0, colour = "grey65", size = 3.2) +
  annotate("text", x = 1, y = -0.5, label = "2015", colour = "grey30",
           size = 5.5, fontface = "bold") +
  annotate("text", x = 2, y = -0.5, label = "2025", colour = "grey30",
           size = 5.5, fontface = "bold") +
  labs(title = "Top Countries by Quality of Life",
       subtitle = "Netherlands climbs 11 places",
       caption = "Grey = entered/exited top 10") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
        plot.subtitle = element_text(colour = "grey50", size = 9, hjust = 0.5),
        plot.caption = element_text(colour = "grey60", size = 7, hjust = 1))
```

## Reference

### Parameters

| Parameter | Default | Description |
|-----------|---------|-------------|
| `method` | `"sigmoid"` | Interpolation method: `"sigmoid"` (logistic S-curve with C1 correction) or `"hermite"` (cubic Hermite smoothstep). See **Interpolation methods** below |
| `smooth` | `8` | Steepness of the sigmoid curve. Higher = sharper S-shape. Only used when `method = "sigmoid"` |
| `width` | `0.8` | Ribbon full width in data units (`geom_bump_ribbon()` only) |
| `n` | `100` | Interpolation points per segment (must be >= 2) |

### Computed variables

| Variable | Description |
|----------|-------------|
| `avg_y` | Mean of y values in the group, inverse-transformed to original data space. Works with `scale_y_reverse()`. |
| `ymin` | Lower ribbon boundary (`geom_bump_ribbon()` only) |
| `ymax` | Upper ribbon boundary (`geom_bump_ribbon()` only) |

### Functions

| Function | Description |
|----------|-------------|
| `geom_bump_ribbon()` | Smooth-curved filled ribbon between rank positions. Supports `method = "sigmoid"` (default) and `method = "hermite"`. Uses `GeomRibbon`. |
| `geom_bump_line()` | Smooth-curved line between rank positions. Same `method` parameter. Uses `GeomPath`. |
| `scale_fill_rank()` | Green-yellow-red gradient scale with `guide = "none"`. Defaults to auto-range from data; pass `limits = c(lo, hi)` for explicit control |
| `theme_bump()` | Dark background theme for rank comparison infographics |

### Interpolation methods

When a group has 3+ time points, adjacent curved segments must be joined smoothly. Both geoms offer two methods via the `method` parameter, each guaranteeing C1 (first-derivative) continuity at every segment join:

**`method = "sigmoid"`** (default) — each segment follows a logistic sigmoid σ(t) = 1/(1+e⁻ᵗ), clamped to exact knot values and with a Hermite-basis derivative correction at interior knots. This preserves the classic ggbump S-curve shape and honours the `smooth` parameter. `smooth` controls steepness: lower values (2–3) produce gentle curves, higher values (12–15) approach step functions.

**`method = "hermite"`** — evaluates a single cubic Hermite spline (`stats::splinefunH()`) with zero slopes at all knots. Visually similar to sigmoid (the smoothstep polynomial 3t²−2t³ closely approximates the logistic) but computed in one pass. The `smooth` parameter is ignored.

### Architecture

```
geom_bump_ribbon(method = "sigmoid" | "hermite"):
  User data (x, y, group)
    → StatBumpRibbon$compute_group()
      → smooth_path(method) for upper + lower edges at y ± width/2
        → "sigmoid": smooth_path_sigmoid() — clamped logistic + Hermite correction
        → "hermite": smooth_path_hermite() — stats::splinefunH() with zero slopes
      → returns data.frame(x, y, ymin, ymax, avg_y)
    → GeomRibbon renders filled area

geom_bump_line(method = "sigmoid" | "hermite"):
  User data (x, y, group)
    → StatBumpLine$compute_group()
      → smooth_path(method) for centerline at y
      → returns data.frame(x, y, avg_y)
    → GeomPath renders stroked path

Both Stats share smooth_path(), the avg_y inverse-transform, and C1 continuity at joins.
```

## Comparison

| Package | What it does | What it lacks |
|---------|-------------|---------------|
| **ggbump** | Sigmoid *lines* via `geom_bump()` | No filled ribbons |
| **ggforce** | Bezier *ribbons* via `geom_diagonal_wide()` | Bezier, not sigmoid curve shape |
| **ggsankey** | Sankey-style ribbon bumps | *Stacked* positioning, not rank-positioned. GitHub-only |
| **ggbumpribbon** | **C1-continuous sigmoid filled ribbons and lines at exact rank positions** | — |

## Dependencies

**Hard:** ggplot2 (>= 3.5.0), cli, rlang, scales — all are already ggplot2 dependencies. The `"hermite"` method uses `stats::splinefunH()` from base R.

## License

MIT
