---
title: "Getting started with ggbumpribbon"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ggbumpribbon}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 96
)
```

```{r setup}
library(ggbumpribbon)
library(ggplot2)
```

## Data format

Both `geom_bump_ribbon()` and `geom_bump_line()` expect a long-format
data frame with columns for position (`x`), rank (`y`), and a grouping
variable. Each group must have at least two x-values.

```{r data}
df <- data.frame(
  x     = rep(1:3, each = 4),
  y     = c(1,2,3,4, 3,1,4,2, 2,4,1,3),
  group = rep(LETTERS[1:4], 3)
)
df
```

## Filled ribbons

`geom_bump_ribbon()` renders each group as a filled area with smooth
sigmoid curves between rank positions. The `fill` aesthetic controls
colour; mapping `after_stat(avg_y)` produces gradient fills based on
mean rank.

```{r ribbon}
ggplot(df, aes(x, y, group = group, fill = after_stat(avg_y))) +
  geom_bump_ribbon(alpha = 0.85) +
  scale_fill_viridis_c(guide = "none") +
  scale_y_reverse() +
  theme_void()
```

### Parameters

The `smooth` parameter controls sigmoid steepness (higher = sharper),
`width` sets the ribbon thickness in data units, and `n` controls
interpolation density.

```{r params-smooth3, fig.width = 6, fig.height = 3}
p_base <- ggplot(df, aes(x, y, group = group, fill = group)) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  scale_y_reverse() + theme_void()

p_base + geom_bump_ribbon(smooth = 3) + ggtitle("smooth = 3")
```

```{r params-smooth15, fig.width = 6, fig.height = 3}
p_base + geom_bump_ribbon(smooth = 15) + ggtitle("smooth = 15")
```

```{r params-width, fig.width = 6, fig.height = 3}
p_base + geom_bump_ribbon(width = 0.4) + ggtitle("width = 0.4")
```

## Lines

`geom_bump_line()` renders the same sigmoid curves as stroked paths.
Map `colour` instead of `fill`.

```{r line}
ggplot(df, aes(x, y, group = group, colour = after_stat(avg_y))) +
  geom_bump_line(linewidth = 1.2) +
  scale_colour_viridis_c(guide = "none") +
  scale_y_reverse() +
  theme_void()
```

## Interpolation methods

Both geoms accept a `method` parameter. The default `"sigmoid"` uses
logistic S-curves with C1-continuous derivative correction at segment
joins. The alternative `"hermite"` uses cubic Hermite interpolation
via `stats::splinefunH()` with zero slopes at knots.

```{r method-sigmoid, fig.width = 6, fig.height = 3}
ggplot(df, aes(x, y, group = group, fill = group)) +
  geom_bump_ribbon(method = "sigmoid") +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  scale_y_reverse() + ggtitle('method = "sigmoid"') + theme_void()
```

```{r method-hermite, fig.width = 6, fig.height = 3}
ggplot(df, aes(x, y, group = group, fill = group)) +
  geom_bump_ribbon(method = "hermite") +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  scale_y_reverse() + ggtitle('method = "hermite"') + theme_void()
```

## Scale and theme

`scale_fill_rank()` provides a green-yellow-red gradient with `guide =
"none"`, defaulting to auto-range from the data. `theme_bump()` gives
a dark background suited to rank comparison infographics.

```{r scale-theme}
ggplot(df, aes(x, y, group = group, fill = after_stat(avg_y))) +
  geom_bump_ribbon() +
  scale_fill_rank() +
  scale_y_reverse() +
  theme_bump()
```

## Labels

Add rank labels with standard ggplot2 layers. A typical pattern:

```{r labels}
lbl_l <- df[df$x == 1, ]
lbl_r <- df[df$x == 3, ]

ggplot(df, aes(x, y, group = group, fill = after_stat(avg_y))) +
  geom_bump_ribbon(alpha = 0.85) +
  scale_fill_viridis_c(guide = "none") +
  scale_y_reverse() +
  scale_x_continuous(limits = c(0.3, 3.7)) +
  geom_text(data = lbl_l, aes(x = 0.92, y = y, label = paste(y, group)),
            inherit.aes = FALSE, hjust = 1, size = 3.5) +
  geom_text(data = lbl_r, aes(x = 3.08, y = y, label = paste(group, y)),
            inherit.aes = FALSE, hjust = 0, size = 3.5) +
  theme_void()
```

## Combining ribbons and lines

Overlay `geom_bump_line()` on top of `geom_bump_ribbon()` for a
bordered appearance:

```{r combined}
ggplot(df, aes(x, y, group = group)) +
  geom_bump_ribbon(aes(fill = group), alpha = 0.6) +
  geom_bump_line(aes(colour = group), linewidth = 0.8) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  scale_colour_brewer(palette = "Set2", guide = "none") +
  scale_y_reverse() +
  theme_void()
```

## Computed variables

`StatBumpRibbon` computes `avg_y` (group mean rank, inverse-transformed
for `scale_y_reverse()` compatibility), `ymin`, and `ymax`.
`StatBumpLine` computes `avg_y` only.

```{r computed}
group_data <- data.frame(x = 1:3, y = c(1, 5, 2))
out <- StatBumpRibbon$compute_group(group_data, NULL, smooth = 8,
                                     n = 50, width = 0.8)
head(out)
```
